The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-04-27 VegOzone.tex
Updates Ane.

Within the scope of our project to assess probable effects of midnight sun conditions on ozone uptake by plants and resultant enhanced susceptibility to ozone inflicted damage, we conducted ambient air ozone concentration measurements at Svanhovd in the growing seasons of 2018/19 in collaboration with NILU. 
In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden at NIBIO Environment Centre Svanhovd. In contrast, no such damage has been reported in 2019.
The respective observed average ozone concentration in 2018/19, however, did not differ significantly by statistical means.

@@

Relevant data for the growing season 2018/19 are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. The hatched areas mark times when no ozone data was recorded. Note, while the downtime during winter was planed, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observations are locate on higher ground than Svanhovd which will result in higher surface ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors. {\bf TODO: Citation!} {\bf TODO: The following has to be confirmed and elaborated on! Ane?} The vegetation at all sites is similar and consists mainly of pine forests, birch, and heath.

Data from before the 1990s is available solely from Svanvik and Karasjok. These data, however, do not follow the high quality standard procedures implemented nowadays and have to be treated with care {\bf TODO: citation NILU ozone report 200x}. Especially, ozone monitors did not undergo a regular re-calibration. This leads likely to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a breaking point in the combined data series denoted as Jergul/Karasjok in the following which would be of a problem if this was the sole data to compare with. Only few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are shown for completion. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas indicate time periods with insufficient quality control as mentioned above.

>>>>

Within the scope of our project to assess probable effects of midnight sun conditions on ozone uptake by plants and resultant enhanced susceptibility to ozone inflicted damage, we conducted ambient air ozone concentration measurements at Svanhovd in the growing seasons of 2018/19 in collaboration with NILU. 
In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden at NIBIO Environment Centre Svanhovd. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.
The respective observed average ozone concentration in 2018/19, however, did not differ significantly by statistical means.

@@

Relevant data for the growing season 2018/19 are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. The hatched areas mark times when no ozone data was recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observations are located on higher ground than Svanhovd which will result in higher surface ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors. {\bf TODO: Citation!} {\bf TODO: The following has to be confirmed and elaborated on! Ane?} The vegetation at all sites is similar and consists mainly of pine forests, birch, and heath.

Data from before the 1990s is available solely from Svanvik and Karasjok. These data, however, do not follow the high quality standard procedures implemented nowadays and have to be treated with care {\bf TODO: citation NILU ozone report 200x}. Especially, ozone monitors did not undergo a regular re-calibration. This probably leads to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a breaking point in the combined data series, denoted as Jergul/Karasjok in the following, which would be of a problem if this was the sole data to compare with. Only few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are shown for completion. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas indicate time periods with insufficient quality control as mentioned above.
